\pkg{partykit}: A Toolkit for Recursive Partytioning

### 2011-09-07 pkg/partykit/inst/doc/partykit.Rnw
renamed vignette file and added introductory WeatherPlay example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@211 255f4249-2263-4417-b1ea-8498779e27a7

%\VignetteIndexEntry{partykit: A Toolkit for Recursive Partytioning}
%\VignetteDepends{partykit}
%\VignetteKeywords{recursive partitioning}
%\VignettePackage{partykit}

@@

%% commands
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

@@

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
   \And Achim Zeileis\\Leopold-Franzens-Universit\"at \\ Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

\Abstract{
  The \pkg{partykit} package 
}
\Keywords{recursive partitioning}

\Address{
  Torsten Hothorn\\
  Institut f\"ur Statistik\\
  Ludwig-Maximilians-Universit\"at M\"unchen\\
  Ludwigstra{\ss}e 33\\
  80539 M\"unchen, Germany\\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://www.stat.uni-muenchen.de/~hothorn/}\\

  Achim Zeileis\\
  Institut f\"ur Statistik \\
  Fakult\"at f\"ur Volkswirtschaft und Statistik \\
  Leopold-Franzens-Universit\"at Innsbruck \\
  Universit\"atsstra{\ss}e 15 \\
  6020 Innsbruck, \"Osterreich \\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://statmath.wu.ac.at/~zeileis/}
}

@@

<<setup, echo = FALSE, results = hide>>=
options(width = 70, SweaveHooks = list(leftpar = 
    function() par(mai = par("mai") * c(1, 1.1, 1, 1))))
require("partykit")
set.seed(290875)
data("iris")
@

\setkeys{Gin}{width=\textwidth}

@@

%%% Aufbau: erstmal Baum zusammenbauen, den plotten (ohne code) und diesen
%%% Baum dann Schritt fuer Schritt zusammensetzen vielleicht:

<<weather-data, echo = FALSE>>=
weather <-
structure(list(outlook = structure(c(1L, 1L, 2L, 3L, 3L, 3L, 
2L, 1L, 1L, 3L, 1L, 2L, 2L, 3L), .Label = c("sunny", "overcast", 
"rainy"), class = "factor"), temperature = c(85, 80, 83, 70, 
68, 65, 64, 72, 69, 75, 75, 72, 81, 71), humidity = c(85, 90, 
86, 96, 80, 70, 65, 95, 70, 80, 70, 90, 75, 91), windy = c(FALSE, 
TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE, TRUE, 
TRUE, FALSE, TRUE), play = structure(c(2L, 2L, 1L, 1L, 1L, 2L, 
1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L), .Label = c("yes", "no"), class = "factor")), .Names = c("outlook", 
"temperature", "humidity", "windy", "play"), row.names = c(NA, 
-14L), class = "data.frame")
@
%%% response: play siehe Witten/Frank im Weka Buch

@@

<<party-1, echo = TRUE>>=
t1 <- party(n1, 
   data = iris,
   fitted = data.frame("(fitted)" = fitted_node(n1, data = iris),
                       "(response)" = iris$Species,
                       check.names = FALSE))
t1
@

\section{My first Tree}

@@

    ### compute terminal node number for each obs.
    fitted <- fitted_node(nodes, data = data)
    ### return rich object
    ret <- party(nodes, 
        data = data,
        fitted = data.frame("(fitted)" = fitted,
                            "(response)" = data[[response]],
                            "(weights)" = weights,
                            check.names = FALSE),
        terms = terms(formula))
    as.constparty(ret)
}
@

@@

\begin{figure}
\begin{center}
<<mytree-5, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt)
@
\caption{Tree. \label{plottree}}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
<<mytree-8, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt4)
@
\caption{Subtree. \label{subtree}}
\end{center}
\end{figure}

>>>>

%\VignetteIndexEntry{partykit: A Toolkit for Recursive Partytioning}
%\VignetteDepends{partykit}
%\VignetteKeywords{recursive partitioning, regression trees, classification trees, decision trees}
%\VignettePackage{partykit}

@@

%% additional commands
\newcommand{\squote}[1]{`{#1}'}
\newcommand{\dquote}[1]{``{#1}''}
\newcommand{\fct}[1]{{\texttt{#1()}\index{#1@\texttt{#1()}}}}
\newcommand{\class}[1]{\dquote{\texttt{#1}}}

%% further commands
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

@@

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
   \And Achim Zeileis\\Universit\"at \\ Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

\Abstract{
  This vignette is intended to be a short technical introduction to the \pkg{partykit} package.
  It is still unfinished but hopefully already helpful to some interested users.

  The purpose of the package is to provide a toolkit with infrastructure for representing,
  summarizing, and visualizing tree-structured regression and classification models.
  Thus, the focus is not on \emph{inferring} such a tree structure from data but
  to \emph{represent} a given tree so that print/plotting and computing predictions
  can be performed in a standardized way. In particular, this unified infrastructure can be
  used for reading/coercing tree models from different sources
  (\pkg{rpart}, \pkg{RWeka}, PMML) yielding objects that share functionality for
  \code{print}/\code{plot}/\code{predict} methods.
}
\Keywords{recursive partitioning, regression trees, classification trees, decision trees}

\Address{
  Torsten Hothorn\\
  Institut f\"ur Statistik\\
  Ludwig-Maximilians-Universit\"at M\"unchen\\
  Ludwigstr.~33\\
  80539 M\"unchen, Germany\\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://www.stat.uni-muenchen.de/~hothorn/}\\

  Achim Zeileis\\
  Department of Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

@@

<<setup, echo = FALSE, results = hide>>=
options(width = 70)
library("partykit")
set.seed(290875)
data("iris")
@

@@

To illustrate how \pkg{partykit} can be used to represent trees,
a simple artificial data set from Witten \& Frank's book
\emph{Data Mining: Practical Machine Learning Tools and Techniques}
is used. It concerns the conditions suitable for playing some
unspecified game.
<<weather-data>>=
data("WeatherPlay", package = "partykit")
WeatherPlay
@
To represent the \code{play} decision based on the corresponding
weather condition variables one could use the tree displayed in
Figure~\ref{weather-plot}. For now, it is ignored how this tree
was \emph{inferred} and it is simply assumed to be given.

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\centering
<<weather-plot0, echo=FALSE, fig=TRUE, height=5, width=7.5>>=
py <- party(
  partynode(1L,
    split = partysplit(1L, index = 1:3),
    kids = list(
      partynode(2L,
        split = partysplit(3L, breaks = 75),
        kids = list(
          partynode(3L, info = "yes"),
          partynode(4L, info = "no"))),
      partynode(5L, info = "yes"),
      partynode(6L,
        split = partysplit(4L, index = 1:2),
        kids = list(
          partynode(7L, info = "yes"),
          partynode(8L, info = "no"))))),
  WeatherPlay)
plot(py)
@
\caption{\label{weather-plot} Decision tree for \code{play} decision based on
  weather conditions in \code{WeatherPlay} data.}
\end{figure}
\setkeys{Gin}{width=\textwidth}

To represent this tree (also known as recursive partition) in \pkg{partykit},
two basic building blocks are used: splits of class \class{partysplit} and
nodes of class \class{partynode}. The resulting recursive partition can then
be associated with a data set in an object of class \class{party}.

\subsubsection*{Splits}

First, we employ the \fct{partysplit} function to creat the three
splits in the ``weather tree'' from Figure~\ref{weather-plot}. The
function takes the following arguments
\begin{Code}
  partysplit(varid, breaks = NULL, index = NULL, ..., info = NULL)
\end{Code}
where \code{varid} is an integer id of the variable used for splitting
e.g., \code{1L} for \code{outlook}, \code{3L} for \code{humidity},
\code{4L} for \code{windy} etc. Then, \code{breaks} and \code{index}
determine which observations are sent to which of the branches, e.g.,
\code{breaks = 75} for the humidity split. Apart from further arguments
not shown above (and just comprised under `\code{...}'), some arbitrary
information can be associated with a \class{partysplit} object by passing
it to the \code{info} argument. The three splits from Figure~\ref{weather-plot}
can then be created via
<<weather-splits>>=
sp_o <- partysplit(1L, index = 1:3)
sp_h <- partysplit(3L, breaks = 75)
sp_w <- partysplit(4L, index = 1:2)
@
For the numeric \code{humidity} variable the \code{breaks} are set while
for the factor variables \code{outlook} and \code{windy} the information
is supplied which of the levels should be associated with which of the
branches of the tree.

\subsubsection*{Nodes}

Second, we use these splits in the creation of the whole decision tree.
In \pkg{partykit} a tree is represented by a \class{partynode} object
which is recursive in that it may have ``kids'' that are again \class{partynode}
objects. These can be created with the function
\begin{Code}
  partynode(id, split = NULL, kids = NULL, ..., info = NULL)
\end{Code}
where \code{id} is an integer identifier of the node number, \code{split}
is a \class{partysplit} object, and \code{kids} is a list of \class{partynode}
objects. Again, there are further arguments not shown (\code{...}) and
arbitrary information can be supplied in \code{info}. The whole tree
from Figure~\ref{weather-plot} can then be created via
<<weather-nodes>>=
pn <- partynode(1L, split = sp_o, kids = list(
  partynode(2L, split = sp_h, kids = list(
    partynode(3L, info = "yes"),
    partynode(4L, info = "no"))),
  partynode(5L, info = "yes"),
  partynode(6L, split = sp_w, kids = list(
    partynode(7L, info = "yes"),
    partynode(8L, info = "no")))))
@
where the previously created \class{partysplit} objects are used as splits
and the nodes are simply numbered from~1 to~8. For the terminal nodes of the
tree there are no \code{kids} and the corresponding \code{play} decision
is stored in the \code{info} argument. Printing the \class{partynode} object
reflects the recursive structure stored.
<<weather-nodes-print>>=
pn
@
However, the displayed information is still rather raw as it has not yet
been associated with the \code{WeatherPlay} data set.

\subsubsection*{Trees (or recursive partitions)}

Hence, in a third step the recursive tree structure stored
in \code{pn} is coupled with the corresponding data in a \class{party}
object.
<<weather-party>>=
py <- party(pn, WeatherPlay)
py
@
In addition to the \class{partynode} and the \class{data.frame},
the function \fct{party} takes several further arguments including
\code{fitted} values, a \code{terms} object, arbitrary additional
\code{info}, and some further arguments (\code{...}):
\begin{Code}
  party(node, data, fitted = NULL, terms = NULL, ..., info = NULL)
\end{Code}

The tree \code{py} can then be easily visualized by
<<weahter-plot, eval=FALSE>>=
plot(py)
@
yielding Figure~\ref{weather-plot}.

\subsubsection*{More}

<<weather-party-print>>=
print(py,
  terminal_panel = function(node) paste(": play=", node$info, sep = ""))
@

%foo_node
%foo_split

%kidids_node(partynode_object)
%kidids_split(partynode_object)

@@

<<party-1, echo = TRUE>>=
t1 <- party(n1, 
  data = iris,
  fitted = data.frame(
    "(fitted)" = fitted_node(n1, data = iris),
    "(response)" = iris$Species,
    check.names = FALSE)
)
t1
@

\section{My first tree}

@@

  ### compute terminal node number for each obs.
  fitted <- fitted_node(nodes, data = data)
  ### return rich object
  ret <- party(nodes, 
    data = data,
    fitted = data.frame(
      "(fitted)" = fitted,
      "(response)" = data[[response]],
      "(weights)" = weights,
      check.names = FALSE),
    terms = terms(formula))
  as.constparty(ret)
}
@

@@

\begin{figure}[t!]
\centering
<<mytree-5, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt)
@
\caption{Tree. \label{plottree}}
\end{figure}

@@

\begin{figure}[t!]
\centering
<<mytree-8, echo = TRUE, fig = TRUE, width = 10, height = 6>>=
plot(aqt4)
@
\caption{Subtree. \label{subtree}}
\end{figure}
